# Reproducible Research: Peer Assessment 1
**Written By: Mike Silva**  

In this analysis I will answer the following questions:  
1. What is mean total number of steps taken per day?  
2. What is the average daily activity pattern?  
3. Are there differences in activity patterns between weekdays and weekends?  

## Loading and preprocessing the data

First we check for the zip archive in the working directory.  If it is not there, it is downloaded.  The data is loaded into the environment.

```{r}
temp <- "activity.zip"
if(!file.exists(temp)){
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
}
data <- read.csv(unz(temp, "activity.csv"))
```

Now that we have the data let's check the structure:

```{r}
str(data)
```

The "date" variable is a factor instead of a date.  We need to change the type.

```{r}
data$date <- as.Date(data$date, "%Y-%m-%d");
```

Checking the structure once again we see:

```{r}
str(data)
```

"Date" is now a date.  Finally we will load the data into a data table object.

```{r}
library(data.table)
data <- as.data.table(data)
```

## What is mean total number of steps taken per day?

To answer this question we must first agregate the data by the date.

```{r}
data.by.date <- data[,sum(steps), by="date"]
setnames(data.by.date, "V1" ,"steps")
```

The following figure illustrates the distribution of the total number of steps taken each day.

```{r}
library(ggplot2)
ggplot(data.by.date, aes(x=steps)) + geom_histogram(colour="black", fill="#fc9272", binwidth=1000) + geom_vline(aes(xintercept=mean(steps, na.rm=T)), color="#de2d26", linetype="dashed", size=1) + ggtitle("Figure 1: Total number of steps taken each day") + xlab("Total Steps per Day") + ylab("Count") + theme(plot.title=element_text(size=rel(1.25), lineheight=2, face="bold"))
```

The verticle red line in the above figure is the mean.  The mean and median are computed:

```{r}
mean(data.by.date$steps, na.rm=T)
median(data.by.date$steps, na.rm=T)
```

The mean and median total number of steps taken per day are **10,766** and **10,765** respectively.

## What is the average daily activity pattern?

To explore this question we first have to aggregate the data by 5-minute intervals.  

``` {r}
data.by.interval <- data[,mean(steps, na.rm=T), by="interval"]
setnames(data.by.interval,"V1","avg.steps")
```

We then created a timeseries plot to visualize the data:

```{r}
ggplot(data=data.by.interval, aes(x=interval, y=avg.steps)) + geom_line() + xlab("5-Minute Interval") + ylab("Average Number of Steps") + ggtitle("Figure 2: Average number of steps across all days by interval") + theme(plot.title=element_text(size=rel(1.25), lineheight=2, face="bold"))
```

We see that there are very few steps util about the 500th 5-minute interval.  The peak is somewhere around 800.

### Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?  

To get a more precise answer we subset the data by interval pulling out the observation with the maximum average steps:

```{r}
data.by.interval[data.by.interval$avg.steps == max(data.by.interval$avg.steps),]
```

Interval **835** had the maximum number of steps (about 206).

## Imputing missing values
When observing the data there are a number of days/intervals where there are missing values (coded as NA).

``` {r}
summary(data)
```

**2,304** observations are missing data.  I will fill in these observations with the mean number of steps for the given 5 minute interval.

```{r}
## Hat tip to http://stackoverflow.com/questions/9322773/how-to-replace-na-with-mean-by-subset-in-r-impute-with-plyr
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
imputed.data <- ddply(data, ~ interval, transform, steps = impute.mean(steps))
```

Now we check to see if there are no NA's

```{r}
summary(imputed.data)
```

Now we want to see a histogram of the total number of steps taken each day.  To do this we need to first aggregate the data by date:

```{r}
imputed.data <- as.data.table(imputed.data)
imputed.data.by.date <- imputed.data[,sum(steps), by="date"]
setnames(imputed.data.by.date, "V1" ,"steps")
```

Now we can plot the histogram:

```{r}
ggplot(imputed.data.by.date, aes(x=steps)) + geom_histogram(colour="black", fill="#9ecae1", binwidth=1000) + geom_vline(aes(xintercept=mean(steps, na.rm=T)), color="#3182bd", linetype="dashed", size=1) + ggtitle("Figure 3: Total number of steps taken each day") + xlab("Total Steps per Day") + ylab("Count") + theme(plot.title=element_text(size=rel(1.25), lineheight=2, face="bold"))
```

### Do these values differ from the first estimates?

As you may recall the mean and median total number of steps taken per day, without imputation, are **10,766** and **10,765** respectively.

```{r}
mean(data.by.date$steps, na.rm=T)
median(data.by.date$steps, na.rm=T)
```

Once the missing data was filled with imputed values the mean and median total number of steps taken per day, are **both 10,766**

```{r}
mean(imputed.data.by.date$steps)
median(imputed.data.by.date$steps)
```

### What is the impact of imputing missing data on the estimates of the total daily number of steps?

By imputing values using the mean, we are reinforced the trends that existed in the data.  Here's another way of looking at it:

```{r}
data.by.date$imputed <- "No"
imputed.data.by.date$imputed <- "Yes"
histogram.data <- rbind(data.by.date, imputed.data.by.date)
ggplot(histogram.data, aes(x=steps, fill=imputed)) + geom_histogram(colour="black", binwidth=1000)+scale_fill_manual(values=c("#fc9272", "#9ecae1"))+ ggtitle("Figure 4: Total number of steps taken each day") + xlab("Total Steps per Day") + ylab("Count") + theme(plot.title=element_text(size=rel(1.25), lineheight=2, face="bold"))
```

## Are there differences in activity patterns between weekdays and weekends?

We will be using the dataset with imputed values for this analysis.  First we determine if the day was a weekday or a weekend:

```{r}
imputed.data$day.of.week <- 'Weekday'
imputed.data[weekdays(imputed.data$date)=="Sunday" | weekdays(imputed.data$date)=="Saturday",4] <- 'Weekend'

table(imputed.data$day.of.week)
```

There are 12,960 weekday and 4,608 weekend observations in the data set.  
To explore this question we first have to aggregate the data by 5-minute intervals.  

``` {r}
imputed.data.by.interval <- imputed.data[,mean(steps, na.rm=T), by="interval,day.of.week"]
setnames(imputed.data.by.interval,"V1","avg.steps")
```

We then created a timeseries plot to visualize the data:

```{r}
ggplot(data=imputed.data.by.interval, aes(x=interval, y=avg.steps, color=day.of.week)) + geom_line() + facet_grid(day.of.week ~ .) + xlab("5-Minute Interval") + ylab("Average Number of Steps") + ggtitle("Figure 5: Average number of steps by interval and day of the week") + theme(plot.title=element_text(size=rel(1.25), lineheight=2, face="bold")) + guides(color=FALSE)
```

A few things stick out when comparing the two lines.  The weekend line increases latter and decreases latter.  This suggests that people sleep in and stay up latter on the weekend.  The weekend also has more steps in the afternoon than the weekday. The weekdays have the most steps.

``` {r}
by(imputed.data.by.interval$avg.steps, imputed.data.by.interval$day.of.week, summary)
```

The weekday average number of steps is about **36** while the weekend is abour **42**.